---
title: "Quantitative Methods for LEL"
subtitle: "Week 11 - p-values"
author: "Elizabeth Pankratz"
institute: "University of Edinburgh"
date: "2023/11/28"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css:
      - ../xaringan-themer.css
      - ../custom.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "../macros.js"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=7, fig.height=5, fig.retina=3,
  out.width = "60%", fig.align = "center",
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
knitr::opts_knit$set(root.dir = here::here())

library(xaringanExtra)
use_xaringan_extra(c("panelset", "tachyons", "freezeframe"))

library(tidyverse)
theme_set(theme_light())
library(brms)
library(extraDistr)
library(ggdist)
library(glue)
library(posterior)

theme_update(text = element_text(size=14))

options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(8, "Dark2"))
options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(8, "Dark2"))
options(show.signif.stars = FALSE)
my_seed <- 8878
set.seed(my_seed)
```


## Hello world

hello

---

layout: false
layout: true

## Do participants in an experiment speed up as the experiment goes on?

---

```{r}
sl_dat <- read_csv('data/sl_2afc_rts.csv')

set.seed(5)
(p_idx_rt <- sl_dat %>% 
    ggplot(aes(x = idx_c, y = logRT)) +
    geom_jitter(alpha = 0.15, width = 0.2) +
    labs(
      x = 'Trial number (centred)',
      y = 'Logged reaction times',
      caption = 'N = 150 participants'
    )
)
```


---


```{r}
set.seed(5)
(p_idx_rt <- p_idx_rt +
  geom_smooth(method = 'lm'))
```


---

layout: false

# Lorem ipsum

abc

---

# Return to the reaction time data

```{r}
set.seed(5)
p_idx_rt
```

---

layout: false
layout: true

# Fit a frequentist linear model using `lm()`

---

```{r echo = TRUE}
rt_lm <- lm(
  logRT ~ idx_c,  # model formula (like we're used to!)
  data = sl_dat   # name the dataset
)
```

--

```{r echo = TRUE}
summary(rt_lm)
```

---

```{r}
cat(capture.output(summary(rt_lm))[9:12], sep = "\n")
```
--

.bg-washed-green.b--dark-green.ba.bw2.br3.shadow-5.ph4.mt2[
.center[
`Pr(>|t|) <2e-16 `

means that

$p$ < 0.0000000000000002 
]
]

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[

**`(Intercept)`:**

- The logged RT at the mean trial index (i.e., halfway through the experiment) is 7.03 (SD = 0.01).
- H0: `(Intercept)` is equal to 0.
  - $p$ < 0.0000000000000002 
  - Can we reject this null hypothesis?
  - Is it *interesting* to reject this null hypothesis?
]

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[

**`idx_c`:**

- For one unit change in speech rate, logged vowel duration changes by –0.22 to -0.26 log-ms, at 95% probability.

]

---
