---
title: "QML - Week 5"
subtitle: "Regression models: the basics"
author: "Stefano Coretta"
format: revealjs
bibliography: references.bib
---

## Biodiversity and linguistic diversity

```{r}
#| label: setup
#| include: false

library(tidyverse)
theme_set(theme_light())
library(brms)
library(bayesplot)
library(biolingR)
data("bii_nlangs")
```

::: {.callout-tip appearance="simple"}
**What is the relationship between a country's intactness of biodiversity and number of languages spoken by that country's people?**
:::

. . .

::: {.callout-note appearance="simple"}
-   Data from the [biolingR](https://github.com/stefanocoretta/biolingR) package.

    -   `bii_nlangs`: combines data from the [Biodiversity Intactness Index](https://doi.org/10.5519/he1eqmg1) data set and data from [Glottolog](https://glottolog.org).

-   **Biodiversity Intactness Index (BII)**

    -   Ratio of the estimated original number of species in undisturbed sites vs their number in adjacent sites with high human activity.

-   **Number of languages** per country

    -   Simply the number of languages per country as listed on Glottolog.
:::

## The data

```{r}
#| label: bii-nlags
#| echo: true

bii_nlangs

```

## Biodiversity Intactness Index

```{r}
#| label: bii-mean

bii_nlangs <- drop_na(bii_nlangs)

bii_nlangs |> 
  ggplot(aes(bii_mean)) +
  geom_density() +
  geom_rug() +
  labs(
    x = "BII (country mean)"
  )
```

## Biodiversity Intactness Index: logit-transformed

```{r}
#| label: bii-mean-logit

bii_nlangs |> 
  ggplot(aes(bii_logit_mean)) +
  geom_density() +
  geom_rug(alpha = 0.25) +
  labs(
    x = "BII (logit country mean)"
  )
```

## Number of languages

```{r}
#| label: n-langs

bii_nlangs |> 
  ggplot(aes(n_langs)) +
  geom_histogram(binwidth = 10) +
  labs(
    x = "Number of languages (per country)",
    subtitle = "Bin-width = 10"
  )
```

## Number of languages: logged

```{r}
#| label: n-langs-log

bii_nlangs |> 
  ggplot(aes(log_n_langs)) +
  geom_density() +
  geom_rug(alpha = 0.1) +
  labs(
    x = "Logged number of languages (per country)"
  )
```

## BII vs number of languages

```{r}
#| label: bii-nlangs

bii_nlangs |> 
  ggplot(aes(bii_logit_mean, log_n_langs)) +
  geom_point(colour = "darkgreen") +
  labs(
    x = "BII (logit)",
    y = "Number of languages (logged)"
  )

```
