---
title: "QML - Week 6"
subtitle: "Regression models: the basics"
author: "Stefano Coretta"
format: revealjs
bibliography: references.bib
---

## 

```{=html}
<iframe allowfullscreen frameborder="0" height="80%" mozallowfullscreen style="min-width: 500px; min-height: 355px" src="https://app.wooclap.com/events/RRYQTLH/questions/6a54dcad1a6fa95d6c99cdc4" width="100%">

</iframe>
```

## Word frequency and reaction times

```{r}
#| label: setup
#| include: false

library(tidyverse)
theme_set(theme_light())
library(brms)
library(bayesplot)
library(ggdist)
```

::: {.callout-tip appearance="simple"}
**What is the relationship between a word's lexical frequency and reaction times in a lexical decision task in Croatian?**
:::

. . .

::: {.callout-note appearance="simple"}
- Data from [Lexical decision times for nouns from the Croatian Psycholinguistic Database](https://data.crossda.hr/dataset.xhtml?persistentId=doi:10.23669/EV3NFJ&version=2.0).

- Lexical decision task (*is it a real Croatian word?*)

  - **Reaction times**.

  - **Word frequency**: counts from the Croatian web Corpus *hrWaC*.
:::

## Which relationship?

```{=html}
<iframe allowfullscreen frameborder="0" height="80%" mozallowfullscreen style="min-width: 500px; min-height: 355px" src="https://app.wooclap.com/events/RRYQTLH/questions/6a423e4fdf37a75258bb186d" width="100%"></iframe>
```

## Reaction times

```{r}
#| label: rts-plot
#| fig-align: center
#| message: false

croat <- read_csv("data/vlasicek2024/croatian-lexdes.csv") |> 
  mutate(
    log_freq = log(word_frequency),
    rt = rt_milliseconds_mean
  )

croat |> 
  ggplot(aes(rt_milliseconds_mean)) +
  geom_density(fill = "darkgreen", alpha = 0.5) +
  geom_rug(colour = "darkorange", alpha = 0.05) +
  labs(
    x = "RT (ms)"
  )

```

## Word frequency and RTs

```{r}
#| label: rt-freq-plot
#| fig-align: center

croat |> 
  ggplot(aes(word_frequency/1000000, rt)) +
  geom_point(colour = "darkorange", alpha = 0.5) +
  labs(
    x = "Lexical frequency (millions)",
    y = "RT (ms)"
  )
```

## Logged word frequency and RTs

```{r}
#| label: rt-logfreq-plot
#| fig-align: center

croat |> 
  ggplot(aes(log_freq, rt)) +
  geom_point(colour = "darkorange", alpha = 0.5) +
  labs(
    x = "Lexical frequency (logged)",
    y = "RT (ms)"
  )
```

## Gaussian model of RT

$$RT \sim Gaussian(\mu, \sigma)$$

. . .

::: {.callout-important appearance="simple"}
But we want to know what happens to RTs depending on the value of lexical frequency...
:::

. . .

::: {.callout-tip appearance="simple"}
Then we let the mean $\mu$ vary by lexical frequency!
:::

$$\begin{aligned}
RT_i & \sim Gaussian(\mu_i, \sigma)\\
\mu_i & = \beta_0 + \beta_1 \cdot logf_i
\end{aligned}$$

## The equation of a line

[$$y = \beta_0 + \beta_1 \cdot x$$]{.t2em}

::: {.callout-note appearance="simple"}
- $\beta_0$ is the line **intercept**: the $y$ value when $x$ is `0` zero.

- $\beta_1$ is the line **slope**: the change in $y$ for each unit-increase of $x$.
:::

```{r}
#| label: int-plot
#| fig-asp: 1
#| fig-align: center
#| eval: false

ggplot() +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linewidth = 1, colour = "purple") +
  geom_abline(intercept = 2, slope = 1, linewidth = 1, colour = "darkorange") +
  annotate("point", x = 0, y = 0, colour = "purple", size = 5) +
  annotate("point", x = 0, y = 2, colour = "darkorange", size = 5) +
  scale_x_continuous(breaks = seq(-6, 6), limits = c(-6, 6)) +
  scale_y_continuous(breaks = seq(-6, 6), limits = c(-6, 6)) +
  labs(
    x = "X", y = "Y"
  ) +
  theme(text = element_text(size = 30))

```

```{r}
#| label: slo-plot
#| fig-asp: 1
#| fig-align: center
#| eval: false

ggplot() +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linewidth = 1, colour = "purple") +
  geom_abline(intercept = 0, slope = 2, linewidth = 1, colour = "darkorange") +
  annotate("segment", x = -Inf, xend = 1, y = 1, colour = "purple", linewidth = 1, linetype = "dotted") +
  annotate("segment", x = -Inf, xend = 1, y = 2, colour = "darkorange", linewidth = 1, linetype = "dotted") +
  annotate("segment", y = -Inf, yend = 2, x = 1, linewidth = 1, linetype = "dotted") +
  annotate("point", x = 1, y = 1, colour = "purple", size = 5) +
  annotate("point", x = 1, y = 2, colour = "darkorange", size = 5) +
  scale_x_continuous(breaks = seq(-6, 6), limits = c(-6, 6)) +
  scale_y_continuous(breaks = seq(-6, 6), limits = c(-6, 6)) +
  labs(
    x = "X", y = "Y"
  ) +
  theme(text = element_text(size = 30))
```

## Regression model

$$\begin{aligned}
RT_i & \sim Gaussian(\mu_i, \sigma)\\
\mu_i & = \beta_0 + \beta_1 \cdot logf_i & \text{[Regression equation]}
\end{aligned}$$

::: {.callout-tip appearance="simple"}
- A **regression model** is a model that uses the equation of a line (the *regression equation*).

- The model **estimates** $\beta_0$ (the intercept), $\beta_1$ (the slope) and $\sigma$ from the data (i.e. the observed $RT$ and $logf$ values).
:::

. . .

::: {.callout-note appearance="simple"}
$\beta_0$, **intercept**

- Mean RT value *when logged frequency is `0` zero* (i.e. when word frequency is 1; `exp(0)` = 1).

$\beta_1$, **slope**

- *Change* in mean RT for each unit increase of log-frequency (when log-frequency goes from $x$ to $x + 1$).
:::

## 

```{=html}
<iframe allowfullscreen frameborder="0" height="80%" mozallowfullscreen style="min-width: 500px; min-height: 355px" src="https://app.wooclap.com/events/RRYQTLH/questions/6a423e4f8cffe6b5916a2ccf" width="100%"></iframe>
```

## Fit model in brms

```{r}
#| label: rt-bm
#| echo: true

rt_bm <- brm(
  rt ~ 1 + log_freq,
  family = gaussian,
  data = croat,
  cores = 4,
  seed = 4125,
  file = "cache/rt_bm"
)
```

## Model results

```{r}
#| label: rt-bm-posteriors
#| echo: true

library(bayesplot)

mcmc_dens(rt_bm, pars = c("b_Intercept", "b_log_freq", "sigma"))
```

## Posterior draws

```{r}
#| laber: rt-bm-draws
#| echo: true

rt_draws <- as_draws_df(rt_bm)
rt_draws
```

## Posterior of $\beta_1$

```{r}
#| label: rt-post-beta1
#| echo: true

rt_draws |> 
  ggplot(aes(b_log_freq)) +
  geom_density(fill = "brown", alpha = 0.5)
```

## Posterior of $\beta_1$ with CrIs

```{r}
#| label: rt-post-beta-1-cris

library(ggdist)

rt_draws |>
  ggplot(aes(x = b_log_freq)) +
  stat_halfeye(
    .width = c(0.5, 0.8, 0.9),
    aes(fill = after_stat(level))
  ) +
  scale_fill_brewer(na.translate = FALSE)
```

## CrIs of $\beta_1$

```{r}
#| label: rt-beta-1-cris
#| echo: true

library(posterior)

# 95%
quantile2(rt_draws$b_log_freq, c(0.025, 0.975)) |> round()
# 80%
quantile2(rt_draws$b_log_freq, c(0.1, 0.9)) |> round()
# 60%
quantile2(rt_draws$b_log_freq, c(0.2, 0.8)) |> round()
```

## Expected values

$$\begin{aligned}
\mu_i & = \beta_0 + \beta_1 \cdot logf_i\\
\mu_{[logf = 4]} & = \beta_0 + \beta_1 \cdot 4\\
\mu_{[logf = 12]} & = \beta_0 + \beta_1 \cdot 12\\
\end{aligned}$$

. . .

```{r}
#| label: epreds
#| echo: true

rt_draws <- rt_draws |> 
  mutate(
    rt_logf_4 = b_Intercept + b_log_freq * 4,
    rt_logf_12 = b_Intercept + b_log_freq * 12
  )

quantile2(rt_draws$rt_logf_4, c(0.1, 0.9)) |> round()
quantile2(rt_draws$rt_logf_12, c(0.1, 0.9)) |> round()
```

## Plotting expected values

```{r}
#| label: rt-bm-cond
#| fig-align: center

p <- plot(
  conditional_effects(rt_bm),
  points = TRUE,
  point_args = list(alpha = 0.1),
  line_args = list(colour = "darkorange"),
  plot = FALSE
)[[1]]

p +
  labs(x = "Lexical frequency (logged)", y = "Expected RT (ms)")
```

## Word frequency and reaction times (bis)

::: {.callout-tip appearance="simple"}
**What is the relationship between a word's lexical frequency and reaction times in a lexical decision task in Croatian?**
:::

. . .

::: {.callout-note appearance="simple"}
- When log-frequency is 0, the mean RTs are between 1084 and 1129 ms at 95% confidence.

- For each unit increase of log-frequency, the mean RTs decrease by 43-48 ms, at 95% confidence.
:::

## Correlation in NOT causation

::: callout-important
### Be careful!

- **Correlation** between two variables: they co-vary, i.e. they show a systematic association (their values tend to vary together in a consistent pattern).

- **Spurious correlations**: two variables can be correlated because of bias from another variable.
:::

. . .

::: {.callout-note appearance="simple"}
1.  Number of plant names in a language vs. biodiversity of the region
    - Languages in biodiverse regions have more words for plants.
    - **Mediator**: cultural reliance on plants.
2.  Language endangerment vs. economic development
    - Higher economic development is associated with greater language endangerment.
    - **Confounder**: colonial history.
3.  Language prestige vs government policy
    - High prestige languages and officially supported languages each attract learners.
    - **Collider**: If you only look at languages with many learners, prestige and policy might appear related even if they're not causally connected.
:::

## But it is if you use causal inference...

::: callout-tip
### Causal inference

- Correlation *can* be interpreted causally if you adopt a **causal inference** approach.

- Learn about it in McElreath's textbook *Statistical Rethinking*.
  Also check [STeW](https://uoelel.github.io/stew.html).
:::
