---
title: "Bar charts"
subtitle: "Learn how to plot bar charts with ggplot2"
format: 
  html:
    css: [webex.css]
    include-after-body: [webex.js]
date: 2024-07-15
---

```{r}
#| label: packages
#| echo: false
#| message: false

library(tidyverse)
library(webexercises)
```

## Bar charts

::: callout-tip
#### Bar charts

**Bar charts** are useful when you are counting things. For example:

-   Number of verbs vs nouns vs adjectives in a corpus.

-   Number of languages by geographic area.

-   Number of correct vs incorrect responses.

The bar chart geometry is `geom_bar()`.
:::

We will first create a plot with counts of the number of languages in `global_south` by their endangerment status and then a plot where we also split the counts by macro-area.

### Number of languages of the Global South by status

To create a bar chart, you can use the `geom_bar()` geometry.

::: callout-warning
#### Bar chart axes

In a simple bar chart, **you only need to specify one axis, the *x*-axis**, in the aesthetics `aes()`.

This is because the counts that are placed on the *y*-axis are calculated by the `geom_bar()` function under the hood.

This quirk is something that confuses many new learners, so make sure you internalise this.
:::

Go ahead and complete the following code to create a bar chart.

```{r}
#| label: global-south
#| echo: false

library(tidyverse)

glot_status <- readRDS("data/coretta2022/glot_status.rds")

global_south <- glot_status |>
  filter(
    Macroarea %in% c("Africa", "Australia", "Papunesia", "South America"),
    status %in% c("threatened", "shifting", "moribund", "nearly_extinct")
  )
```

```{r}
#| label: status-bar-ex
#| eval: false

global_south |>
  ggplot(aes(x = status)) +
  ...
```

Note how we're using `|>` to pipe the `glot_status` data frame into the `ggplot()` function. This works because `ggplot()`'s first argument is the data, and piping is a different way of providing the first argument to a function.

As mentioned above, the counting for the *y*-axis is done automatically. R looks in the `status` column and counts how many times each value in the column occurs in the data frame.

If you did things correctly, you should get the following plot.

```{r}
#| label: fig-status-bar
#| fig-cap: "Number of languages by endangerment status."
#| echo: false

global_south |>
  ggplot(aes(x = status)) +
  geom_bar()
```

The *x*-axis is now `status` and the *y*-axis corresponds to the number of languages by status (`count`). As mentioned above, `count` is calculated under the hood for you (you will learn how to count levels with `count()` later in the course).

You could write a description of the plot that goes like this:

> The number of languages in the Global South by endangered status is shown as a bar chart in @fig-status-bar. Among the languages that are endangered, the majority are threatened or shifting.

What if we want to show the number of languages by endangerment status within each of the macro-areas that make up the Global South? Easy! You can make a stacked bar chart. This requires we mutate the data first.

## Stacked bar charts

A special type of bar charts are the so-called stacked bar charts.

::: callout-tip
#### Stacked bar chart

A **stacked bar chart** is a bar chart in which each contains a "stack" of shorter bars, each indicating the counts of some sub-groups.

This type of plot is useful to show how counts of something vary depending on some other grouping (in other words, when you want to count the occurrences of a categorical variable based on another categorical variable). For example:

-   Number of languages by endangerment status, grouped by geographic area.

-   Number of infants by head-turning preference, grouped by first language.

-   Number of past vs non-past verbs, grouped by verb class.
:::

To create a stacked bar chart, you just need to add a new aesthetic mapping to `aes()`: `fill`. The `fill` aesthetic lets you fill bars or areas with different colours depending on the values of a specified column.

Let's make a plot on language endangerment by macro-area.

Complete the following code by specifying that `fill` should be based on `status`.

```{r}
#| label: status-stacked-ex
#| eval: false

global_south |>
  ggplot(aes(x = Macroarea, ...)) +
  geom_bar()
```

You should get the following.

```{r}
#| label: fig-status-stacked
#| fig-cap: "Number of languages by macro-area and endangerment status."
#| echo: false

global_south |>
  ggplot(aes(x = Macroarea, fill = status)) +
  geom_bar()
```

A write-up example:

> @fig-status-stacked shows the number of languages by geographic macro-area, subdivided by endangerment status. Africa, Eurasia and Papunesia have substantially more languages than the other areas.

::: callout-note
#### Quiz 4

What is wrong in the following code?

``` r
gestures |>
  ggplot(aes(x = status), fill = Macroarea) +
  geom_bar()
```
:::

## Filled stacked bar charts

In the plot above it is difficult to assess whether different macro-areas have different proportions of endangerment. This is because the overall number of languages per area differs between areas.

A solution to this is to plot **proportions** instead of raw counts.

You could calculate the proportions yourself, but there is a quicker way: using the `position` argument in `geom_bar()`.

You can plot proportions instead of counts by setting `position = "fill"` inside `geom_bar()`, like so:

```{r}
#| label: fig-status-stacked-2
#| fig-cap: "Proportion of languages by macro-area and endangerment status."

global_south |>
ggplot(aes(x = Macroarea, fill = status)) +
  geom_bar(position = "fill")
```

The plot now shows proportions of languages by endangerment status for each area separately.

Note that the *y*-axis label is still "count" but should be "proportion". Use `labs()` to change the axes labels and the legend name.

```{r}
#| label: status-stacked-3-ex
#| eval: false

global_south |>
ggplot(aes(x = Macroarea, fill = status)) +
  geom_bar(position = "fill") +
  labs(
    ...
  )
```

::: {.callout-tip collapse="true"}
#### Hint

If to change the name of the `colour` legend, you use the `colour` argument in `labs()`, guess which argument you should use for `fill`?
:::

You should get this.

```{r}
#| label: fig-status-stacked-3
#| fig-cap: "Proportion of languages by macro-area and endangerment status."
#| echo: false

global_south |>
ggplot(aes(x = Macroarea, fill = status)) +
  geom_bar(position = "fill") +
  labs(
    x = "Macro-area",
    y = "Proportion",
    fill = "Endangerment"
  )
```

With this plot it is easier to see that different areas have different proportions of endangerment. In writing:

> @fig-status-stacked-3 shows proportions of languages by endangerment status for each macro-area. Australia, South and North America have a substantially higher proportion of extinct languages than the other areas. These areas also have a higher proportion of near extinct languages. On the other hand, Africa has the greatest proportion of non-endangered languages followed by Papunesia and Eurasia, while North and South America are among the areas with the lower proportion, together with Australia which has the lowest.

## Faceting and panels

Sometimes we might want to separate the data into separate panels within the same plot.

We can achieve that easily using **faceting**.

### Polite again

Let's reproduce [this plot](tutorial-w04.qmd#lets-add-geometries) from Week 4, but this time let's spice things up.

This is the plot you made in Week 4. Try and reproduce it by writing the code yourself (you also have to read in the data!).

```{r}
#| label: fig-polite
#| fig-cap: "Scatter plot of mean f0 and H1-H2 difference."
#| echo: false
#| message: false
#| warning: false

polite <- read_csv("data/winter2012/polite.csv")

ggplot(
  data = polite,
  mapping = aes(x = f0mn, y = H1H2, colour = gender)
) +
  geom_point()
```

### Does being a music student matter?

That looks great, but we want to know if being a music student has an effect on the relationship of `f0mn` and `H1H2`.

In the plot above, the aesthetics mappings are the following:

-   `f0mn` on the *x*-axis.
-   `H1H2` on the *y*-axis.
-   `gender` as colour.

How can we separate data further depending on whether the participant is a music student or not (`musicstudent`)?

We can create panels using `facet_grid()`. This function takes lists of variables to specify panels in rows and/or columns.

#### Faceting

**Faceting** a plot allows to split the plot into multiple panels, arranged in rows and columns, based on one or more variables.

To facet a plot, use the `facet_grid()` function.

The syntax is a bit strange. You can specify rows of panels with the `rows` argument and columns of panels with `cols` argument, but you have to include column names inside `vars()`, like this:

```{r}
#| label: fig-polite-facet
#| fig-cap: "Scatter plot of mean f0 and H1-H2 difference in non-music students (left) vs music students (right)."
#| warning: false

polite |>
  ggplot(aes(f0mn, H1H2, colour = gender)) +
  geom_point() +
  facet_grid(cols = vars(musicstudent)) +
  labs(
    x = "Mean f0 (Hz)",
    y = "H1-H2 difference (dB)",
    colour = "Gender"
  )
```

You could write a description of this plot like this:

> @fig-polite-facet shows mean f0 and H1-H2 difference as a scatter plot. The two panels indicate whether the participant was a student of music. Within each panel, the participant's gender is represented by colour (red for female and blue for male). Male participants tend to have higher H1-H2 differences and lower mean f0 than females. From the plot it can also be seen that there is greater variability in H1-H2 difference in female music students compared to female non-music participants. Within each group of gender by music student there does not seem to be any specific relation between mean f0 and H1-H2 difference.

The `polite` data also has a column `attitude` with values `inf` for informal and `pol` for polite. Subjects were asked to speak either as if they were talking to a friend (`inf` attitude) or to someone with a higher status (`pol` attitude).

Recreate the last plot, this time faceting also by `attitude`. Use the `rows` column to create two separate rows for each value of `attitude`.

```{r}
#| label: fig-polite-facet-2
#| eval: false

polite |>
  ggplot(aes(f0mn, H1H2, colour = gender)) +
  geom_point() +
  facet_grid(cols = vars(musicstudent), rows = ...)
```

Now write a description of the plot.
