---
title: "Data types"
description: "Learn about types of data and how to import data in R"
author: "Stefano Coretta"
date: 2023-06-28
editor: visual
format: 
  html:
    css: [webex.css]
    include-after-body: [webex.js]
---

```{r}
#| label: packages
#| echo: false
#| message: false

library(tidyverse)
library(webexercises)
```

::: callout-warning
#### Pre-requisites

-   [posts/intro-quarto.qmd](posts/intro-quarto.qmd)
:::

## Tabular data

::: callout-tip
#### Tabular data

Tabular data is data that has a form of a table: i.e. values structured in columns and rows.
:::

Most of the data we will be using in this course will be tabular and the files will be in the `.csv` format.

The **comma separated values format** (`.csv`) is the best format to save data in because it is basically a plain text file, it's quick to parse, and can be opened and edited with any software (plus, it’s not a proprietary format like `.docx` or `.xlsx`---these formats are specific to particular software).

This is what a `.csv` file looks like when you open it in a text editor (showing only the first few lines).

``` txt
Group,ID,List,Target,ACC,RT,logRT,Critical_Filler,Word_Nonword,Relation_type,Branching
L1,L1_01,A,banoshment,1,423,6.0474,Filler,Nonword,Phonological,NA
L1,L1_01,A,unawareness,1,603,6.4019,Critical,Word,Unrelated,Left
L1,L1_01,A,unholiness,1,739,6.6053,Critical,Word,Constituent,Left
L1,L1_01,A,bictimize,1,510,6.2344,Filler,Nonword,Phonological,NA
```

The file contains tabular data (data that is structured as columns and rows, like a spreadsheet).

To separate the values of each column, a `.csv` file uses a comma `,` (hence the name "comma separated values") to separate the values in every row.

The first line of the file indicates the names of the columns of the table:

``` txt
Group,ID,List,Target,ACC,RT,logRT,Critical_Filler,Word_Nonword,Relation_type,Branching
```

There are 11 columns. The rest of the rows is the data, i.e. the values of each column separated by commas.

``` txt
L1,L1_01,A,banoshment,1,423,6.0474,Filler,Nonword,Phonological,NA
L1,L1_01,A,unawareness,1,603,6.4019,Critical,Word,Unrelated,Left
L1,L1_01,A,unholiness,1,739,6.6053,Critical,Word,Constituent,Left
L1,L1_01,A,bictimize,1,510,6.2344,Filler,Nonword,Phonological,NA
```

This might look a bit confusing, but you will see later that, after importing this type of file, you can view it as a nice spreadsheet (as you would in Excel).

Another common type of tabular data file is **spreadsheets**, like spreadsheets created by Microsoft Excel or Apple Numbers. These are all proprietary formats that require you to have the software that were created with if you want to modify them.

Portability and openness are important aspects of conducting ethical research, so that using open and non-proprietary file types makes your research more accessible and doesn't privilege those who have access to specific software (remember, R is free!).

There are also variations of the comma separated values type, like **tab separated values** files (`.tsv`, which uses tab characters instead of commas) and **fixed-width** files (usually `.txt`, where columns are separated by as many white spaces as needed so that the columns align).

### Non-tabular data

Of course, R can import also data that is not tabular, like map data and complex hierarchical data.

We will dip our toes into map data at the end of course, but virtually all of the data we will use will be tabular, just because that's the format you need to do data visualisation and analyses.

### `.rds` files

R has a special way of saving data: `.rds` files.

`.rds` files allow you to save an R object to a file on your computer, so that you can read that file in when you need it.

A common use for `.rds` files is to save tabular data that you have processed so that it can be readily used in many different scripts or even by other people.

In the following sections you will learn how to import (aka read) three types of data: `.csv`, Excel and `.rds` files.

## Download the data files

Throughout the course we will be using data files that come from linguistic research. You should download now the data files according to the following instructions

**Please, follow these instructions carefully.**

1.  Download the zip archive with all the data by right-clicking on the following link and download the file: [data.zip](https://uoelel.github.io/qm-data/data.zip).

2.  Unzip the zip file to extract the contents. (If you don't know how to do this, ask one of the tutors to help you!)

3.  Create a folder called `data/` (the slash is there just to remind you that it's a folder, but you don't have to include it in the name) in the Quarto project you are using for the course.

    1.  To create a folder, go to the Files tab of the bottom-right panel in RStudio.

    2.  Make sure you are viewing the project's main folder.

    3.  Click on the `New Folder` button, enter "data" in the text box and click `OK`

4.  Move the contents of the `data.zip` archive into the `data/` folder.

    1.  Open a Finder or File Explorer window.

    2.  Navigate to the folder where you have extracted the zip file (it will very likely be the `Downloads/` folder).

    3.  Copy the contents of the zip file.

    4.  In Finder or File Explorer, navigate to the Quarto project folder, then the `data/` folder, and paste the contents in there. (You can also drag and drop if you prefer.)

The rest of the tutorial will assume that you have created a folder called `data/` in the Quarto project folder and that the files you downloaded are in that folder. The data folder should like something like this:

```         
data/
└── cameron2020/
    └── gestures.csv
└── coretta2018/
    └── formants.csv
    └── token-measures.csv
└── ...
```

I recommend that you start being very organised with your files in other projects from now on, whether it's for this course or your dissertation or else. I also suggest to avoid overly nested structures (for example, avoid having one folder for each week for this course. Rather, save all data files in the `data/` folder).

::: {.callout-note collapse="true"}
#### Organising your files

The Open Science Framework has the following recommendations that apply very well to any type of research project.

-   Use **one folder** per project. This will also be your RStudio project folder.

-   Separate **raw data** from **derived data**.

-   Separate **code** from **data**.

-   Make **raw data read-only**.

To learn more about this, check the OSF page [Organising files](https://help.osf.io/article/147-organizing-files).

In brief, what these recommendations mean is that you want a folder for your research project/course/else, and inside the folder two folders: one for data and one for code.

The `data/` folder could further contain `raw/` for raw data (data that should not be lost or changed, for example collected data or annotations) and `derived/` for data that derives from the raw data, for example through automated data processing.

I usually also have a separate folder called `figs/` or `img/` where I save plots. Of course which folders you will have it's ultimately up to you and needs will vary depending on the project and field!
:::

## Import `.csv` files

Let's start with data from this paper: Song et al. 2020. *Second language users exhibit shallow morphological processing*. DOI: [10.1017/S0272263120000170](https://doi.org/10.1017/S0272263120000170).

The study consisted of a lexical decision task in which participants were first shown a prime, followed by a target word for which they had to indicate whether it was a real word or a nonce word.

The prime word belonged to one of three possible groups (`Relation_type` in the data) each of which refers to the morphological relation of the prime and the target word:

-   `Unrelated`: for example, *prolong* (assuming *unkindness* as target, \[\[un-kind\]-ness\]).

-   `Constituent`: *unkind*.

-   `NonConstituent`: *kindness*.

### The tidyverse packages

Importing `.csv` files is very easy. You can use the `read_csv()` function from a collection of R packages known as the [tidyverse](https://www.tidyverse.org).

To import data in R we will use the `read_csv()` function from the [readr](https://readr.tidyverse.org) package, one of the tidyverse packages.

Installing the tidyverse packages is easy: you just need to install the tidyverse package and that will take care of installing the most important packages in the collection (called the ["core" tidyverse packages](https://www.tidyverse.org/packages/)).

Go ahead and install the tidyverse from the Packages tab.

### `read_csv()`

::: callout-warning
#### Did you open the RStudio Quarto project?

Before moving on, make sure that you have opened the RStudio Quarto project correctly (see warning at the top of the tutorial)
:::

The `read_csv()` function from the readr package only requires you to specify the file path as a string (remember, strings are quoted between `" "`, for example `"year_data.txt"`). On my computer, the file path of `song2020/shallow.csv` is `/Users/ste/dal/data/song2020/shallow.csv`, but on your computer the file path will be different, of course.

Also, note that it is not enough to use the `read_csv()` function. You also must assign the output of the `read_csv()` function (i.e. the data we are reading) to a variable, using the assignment arrow `<-`, just like we were assigning values to variables in the previous weeks.

And since the `read_csv()` is a function from the tidyverse, you first need to attach the tidyverse packages with `library(tidyverse)` (remember, you need to attach packages **only once** per session). This will attach the core tidyverse packages, including readr. Of course, you can also attach the individual packages directly: `library(readr)`. If you use `library(tidyverse)` there is no need to attach individual tidyverse packages.

**Before reading the data, create a new R script named `tutorial_w03.R` and save it in the `code/` folder of your Quarto project.**

Generally, you start the script with calls to `library()` to load all the packages you need for the script.

Now we only need one package, tidyverse, but in most cases you will need more than one! The best practice is to attach all of packages first, in the top of your script. Please, get in the habit of doing this from now, so that you can keep your scripts tidy and pretty!

::: callout-warning
Please, don't include `install.packages()` in your R scripts!

Remember, you only have to install a package once, and you can just type it in the Console.

But **DO** include `library()` in your scripts.
:::

At the top of `tutorial_w03.R`, write the following lines of code. Then run the code.

```{r}
#| label: read-shallow

library(tidyverse)

shallow <- read_csv("./data/song2020/shallow.csv")
```

If you look at the `Environment` tab, you will see `song2020` under `Data`.

::: callout-note
#### Data frames and tibbles

In R, a data table is called a **data frame**.

**Tibbles** are special data frame created with the read functions from the tidyverse. If you are curious about the difference, check [this page](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html#:~:text=There%20are%20three%20key%20differences,%2C%20subsetting%2C%20and%20recycling%20rules.).

In this course's tutorials, "data frame" and "tibble" will be used interchangeably (since we are using the read functions from the tidyverse, all resulting data frames will be tibbles).
:::

But wait, what is that `"./data/song2020/shallow.csv"`? That's a **relative path**. We briefly mentioned relative paths above, but let's understand the details now. You will be able to view the data soon.

### Relative paths

::: callout-tip
#### Relative path

A **relative path** is a file path that is relative to a folder. The folder the path starts at is represented by `./`.
:::

When you are using R scripts in Quarto projects, the `./` folder paths are relative to is the project folder! This is true whichever the name of the folder/project and whichever it's location on your computer.

For example, if your project it's called `awesome_proj` and it's in `Downloads/stuff/`, then if you write `./data/results.csv` you really mean `Downloads/stuff/awesome_proj/data/results.csv`!

**How does R know the path is relative to the project folder?**

That is because when working with Quarto projects, all relative paths are relative to the project folder (i.e. the folder with the `.Rproj` file)!

The folder which relative paths are relative to is called the **working directory** (directory is just another way of saying folder).

::: callout-tip
#### Working directory

The **working directory** is the folder which relative paths are relative to.

When using Quarto projects, the working directory is the project folder.
:::

The code `read_csv("./data/song2020/shallow.csv")` above will work because you are using a Quarto project and inside the project folder there is a folder called `data/` and in it there's the `song2020/shallow.csv` file.

So from now on I encourage you to use Quarto projects, R scripts and relative paths always!

The benefit of doing so is that, if you move your project or rename it, or if you share the project with somebody, all the paths will just work because they are relative!

::: callout-tip
#### Get the working directory

You can get the current working directory with the `getwd()` command.

Run it now in the Console! Is the returned path the project folder path?

If not, it might be that you are not working from a Quarto project. Check the top-right corner of RStudio: is the project name in there or do you see `Project (none)?`

If it's the latter, you are not in a Quarto project, but you are running R from somewhere else (meaning, the working directory is somewhere else). If so, close RStudio and open the project.
:::

### View the data

Now we can finally view the data.

The easiest way is to click on the name of the data listed in the `Environment` tab, in the top-right panel of RStudio.

You will see a nicely formatted table, as you would in a programme like Excel.

Data tables in R (i.e. tabular, spread-sheet like data) are called **data frames** or **tibbles**.[^1]

[^1]: A **tibble** is a special data frame. We will learn more about tibbles in the following weeks.

The `shallow` data frame contains 11 columns (called variables in the `Environment` tab). The 11 columns are the following:

-   `Group`: `L1` vs `L2` speakers of English.
-   `ID`: Subject unique ID.
-   `List`: Word list (A to F).
-   `Target`: Target word in the lexical decision trial.
-   `ACC`: Lexical decision response accuracy (`0` incorrect response, `1` correct response).
-   `RT`: Reaction times of response in milliseconds.
-   `logRT`: Logged reaction times.
-   `Critical_Filler`: Whether the trial was a `filler` or `critical`.
-   `Word_Nonword`: Whether the Target was a real `Word` or a `Nonword`.
-   `Relation_type`: The type of relation between prime and target word (`Unrelated`, `NonCostituent`, `Constituent`, `Phonological`).
-   `Branching`: Constituent syntactic branching, `Left` and `Right` (shout out to Charlie Puth).

::: callout-note
#### Quiz 3

```{r}
#| label: quiz-3
#| results: asis
#| echo: false

opts_3 <- c(
   "11",
   "650",
   answer = "6500"
)

cat("How many rows does `shallow` have?", longmcq(opts_3))
```
:::

## Import Excel sheets

To read an Excel file we need first to attach the [readxl](https://readxl.tidyverse.org/index.html) package. It should already be installed, because it comes with the tidyverse. If not, then install it.

```{r}
#| label: readxl

library(readxl)

```

Then we can use the `read_excel()` function. Let's read the file.

```{r}
#| label: los2023

relatives <- read_excel("./data/los2023/relatives.xlsx")

```

Now you can view the tibble `los2023`.

Note that if the Excel file has more than one sheet, you can specify the sheet number when reading the file (the default is `sheet = 1`).

```{r}
#| label: los2023-2

relatives_2 <- read_excel("./data/los2023/relatives.xlsx", sheet = 2)
```

The second sheet in `los2023/relatives.xlx` contains the description of the columns in the first sheet.

## Import `.rds` files

Another useful type of data files is a file type specifically designed for r: `.rds` files.

Usually, each `.rds` file contains one R object, like one tibble.

You can read `.rds` files with the `readRDS()` function.

```{r}
#| label: glot-status

glot_status <- readRDS("./data/coretta2022/glot_status.rds")
```

As always, you need to assign the output of the function to a variable, here `glot_status`.

::: callout-tip
#### `.rds` files

`.rds` files are a type of R file which can store any R object and save it on disk.

R objects can be saved to an `.rds` file with the `saveRDS()` function and they can be read with the `readRDS()` function.
:::

View the `glot_status` tibble now.

It is also very easy to save a tibble to an `.rds` file with the `saveRDS()` function.

For example:

```{r}
#| label: save-rds
#| eval: false

saveRDS(shallow, "./data/song2020/shallow.rds")
```

The first argument is the name of the tibble object and the second argument is the file path to save the object to.

## Practice

::: {.callout-note collapse="true"}
#### Practice 1

Read the following files in R, making sure you use the right `read_*()` function.

-   [koppensteiner2016/takete_maluma.txt](../data/koppensteiner2016/takete_maluma.txt) (a tab separated file)

-   [pankratz2021/si.csv](../data/pankratz2021/si.csv)

-   Go to <https://datashare.ed.ac.uk/handle/10283/4006> and download the file `conflict_data_.xlsx`. Read both sheets ("conflict_data2" and "demographics"). Any issues? (I suggest looking at the spread sheet in Excel if it helps).
:::

## Summary

::: {.callout-note appearance="minimal"}
-   You can import tabular data in R with the `read_*()` functions from the tidyverse package readr.

-   You can view data in RStudio as spreadsheets.
:::
