---
title: "Introduction to regression models"
description: "Learn the basics of regression models, a powerful and flexible, yet simple, way of modelling data"
author: "Stefano Coretta"
date: 2024-06-03
---

```{r}
#| label: setup

```

## Regression to the mean

**Regression models** (also called linear models or linear regression models) are a type of statistical model.

In the most general sense, regression models can be used to estimate the underlying process that generates some data. Somewhat more specifically, regression models use the given data to estimate the set of parameters of the equation that is believed to have generated the data.

::: callout-note
### Why "regression"?

...
:::

The formula of an example of simplest regression model is the following:

$$
y \sim Gaussian(\mu, \sigma)
$$

-   $y$ is the data

-   $\sim$ means "is distributed according to", or "is generated by".

-   $Gaussian(\mu, \sigma)$ is a Gaussian probability distribution with the parameters $\mu$ (the mean) and $\sigma$ (the standard deviation).

## Simulating Gaussian data

Let's move onto an example of how we can use regression models to estimate the mean and standard deviation of data that is distributed according to a Gaussian probability.

Alas, it is very difficult to find in nature data that is truly Gaussian, so we will simulate some.

To make things a little bit more worldly, we will simulate data of human adult height.

```{r}
set.seed(62854)
height <- tibble(
  h = round(rnorm(200, 165, 8))
)
```

```{r}
library(tidyverse)
theme_set(theme_light())
```

```{r}
height |> 
  ggplot(aes(h)) +
  geom_density(fill = "darkgreen", alpha = 0.2) +
  geom_rug() +
  geom_vline(aes(xintercept = mean(h)), colour = "purple", size = 1)
```

```{r}
library(brms)
```

```{r}
h_1 <- brm(
  h ~ 1,
  data = height,
  chain = 1
)
```

We can now inspect the summary of the model.

```{r}
summary(h_1)
```
