---
title: "Density plots"
subtitle: "Learn how to visualise continuous variables with density plots"
format: 
  html:
    css: [webex.css]
    include-after-body: [webex.js]
date: 2024-09-13
---

```{r}
#| label: packages
#| echo: false
#| message: false

library(tidyverse)
library(webexercises)
```

::: callout-warning
#### Prerequisites

- [Plotting basics in R](plotting-basics.qmd)
- [Mutate data](wrangle-mutate.qmd)
:::

## Density plots

::: callout-tip
#### Density plots

**Density plots** show the distribution (i.e. the probability density) of the values of a continuous variable**.**

They are created with `geom_density()`.
:::

Let's plot the VOT data from `alb_vot`.

```{r}
#| label: alb-vot

alb_vot <- read_csv("data/coretta2021/alb-vot.csv") |> 
  mutate(
    # Multiply by 1000 to get ms from s
    vot = (voi_onset - release) * 1000
  )

alb_vot
```


VOT is a numeric continuous variable so density plots are appropriate.

To plot the probability density of a continuous variable, you can use the `density` geometry. Remember, all geometry functions start with `geom_`.

Fill in the ... in the following code to create a density plot of VOT values in `alb_vot`.

```{r}
#| label: vot-dens-ex
#| eval: false

alb_vot %>%
  ggplot(aes(x = vot)) +
  ...
```

Note that to create a density plot, you only need to specify the `x`-axis. The `y`-axis is the probability density, which is automatically calculated (a bit like counts in bar charts, remember?).

This is what the plot should look like.

```{r}
#| label: vot-dens
#| echo: false
#| warning: false

alb_vot %>%
  ggplot(aes(x = vot)) +
  geom_density()
```


### Make things cosy with a rug

The density line shows you a smoothed representation of the data distribution over the VOT values, but you might also want to see the raw data.

You can do so by adding the `rug` geometry. Go ahead and add a rug...

```{r}
#| label: vot-dens-rug-ex
#| eval: false

alb_vot %>%
  ggplot(aes(vot)) +
  geom_density() +
  ...
```

You should get the following:

```{r}
#| label: vot-dens-rug
#| echo: false

alb_vot %>%
  ggplot(aes(vot)) +
  geom_density() +
  geom_rug()
```

Nice huh?

::: callout-tip
#### Rug

Raw data can be shown with a **rug**, i.e. ticks on the axes that mark where the data is.

You can add a rug with `geom_rug()`.
:::

::: callout-note
#### Quiz 1

What can you notice about the distribution of VOT values?

::: {.callout-tip collapse="true"}
#### Hint

Are there multiple peaks in the distribution?
:::
:::

::: callout-note

#### Next

- [Faceting plots](plot-faceting.qmd)
- [Advanced plotting](plot-advanced.qmd)

:::
