---
title: "Faceting plots"
subtitle: "Learn how to split plots into panels with faceting"
format: 
  html:
    css: [webex.css]
    include-after-body: [webex.js]
date: 2024-09-13
---

```{r}
#| label: packages
#| echo: false
#| message: false

library(tidyverse)
library(webexercises)
```

::: callout-warning
#### Prerequisites

- [Plotting basics in R](plotting-basics.qmd)
:::

## Faceting and panels

Sometimes we might want to separate the data into separate panels within the same plot.

We can achieve that easily using **faceting**.

## Politeness and speech

Let's revisit the plots from [Plotting basics in R](plotting-basics.qmd). We will use the `polite` data again.

This is the plot you previously made. Try and reproduce it by writing the code yourself (you also have to read in the data!).

```{r}
#| label: fig-polite
#| fig-cap: "Scatter plot of mean f0 and H1-H2 difference."
#| echo: false
#| message: false
#| warning: false

polite <- read_csv("data/winter2012/polite.csv")

ggplot(
  data = polite,
  mapping = aes(x = f0mn, y = H1H2, colour = gender)
) +
  geom_point()
```

### Does being a music student matter?

That looks great, but we want to know if being a music student has an effect on the relationship of `f0mn` and `H1H2`.

In the plot above, the aesthetics mappings are the following:

-   `f0mn` on the *x*-axis.
-   `H1H2` on the *y*-axis.
-   `gender` as colour.

How can we separate data further depending on whether the participant is a music student or not (`musicstudent`)?

We can create panels using `facet_grid()`. This function takes lists of variables to specify panels in rows and/or columns.

#### Faceting

**Faceting** a plot allows to split the plot into multiple panels, arranged in rows and columns, based on one or more variables.

To facet a plot, use the `facet_grid()` function.

The syntax is a bit strange. You can specify rows of panels with the `rows` argument and columns of panels with `cols` argument, but you have to include column names inside `vars()`, like this:

```{r}
#| label: fig-polite-facet
#| fig-cap: "Scatter plot of mean f0 and H1-H2 difference in non-music students (left) vs music students (right)."
#| warning: false

polite |>
  ggplot(aes(f0mn, H1H2, colour = gender)) +
  geom_point() +
  facet_grid(cols = vars(musicstudent)) +
  labs(
    x = "Mean f0 (Hz)",
    y = "H1-H2 difference (dB)",
    colour = "Gender"
  )
```

You could write a description of this plot like this:

> @fig-polite-facet shows mean f0 and H1-H2 difference as a scatter plot. The two panels indicate whether the participant was a student of music. Within each panel, the participant's gender is represented by colour (red for female and blue for male). Male participants tend to have higher H1-H2 differences and lower mean f0 than females. From the plot it can also be seen that there is greater variability in H1-H2 difference in female music students compared to female non-music participants. Within each group of gender by music student there does not seem to be any specific relation between mean f0 and H1-H2 difference.

The `polite` data also has a column `attitude` with values `inf` for informal and `pol` for polite. Subjects were asked to speak either as if they were talking to a friend (`inf` attitude) or to someone with a higher status (`pol` attitude).

Recreate the last plot, this time faceting also by `attitude`. Use the `rows` column to create two separate rows for each value of `attitude`.

```{r}
#| label: fig-polite-facet-2
#| eval: false

polite |>
  ggplot(aes(f0mn, H1H2, colour = gender)) +
  geom_point() +
  facet_grid(cols = vars(musicstudent), rows = ...)
```

Now write a description of the plot.
