---
title: "Regression models: interactions between categorical predictors"
description: "Add and interpret interactions of two categorical predictors"
author: "Stefano Coretta"
date: 2024-08-015
execute: 
  freeze: auto
format: 
  html:
    css: [webex.css]
    include-after-body: [webex.js]
    embed-resources: true 
---

```{r}
#| label: setup
#| include: false
library(webexercises)
library(tidyverse)
library(brms)
```

::: callout-warning
## Prerequisites

- [Regression models: interactions](regression-interactions.qmd).
:::

## Introduction

Here you will learn how to include interactions between categorical predictors and how to interpret the output. Relative to categorical-numeric interactions, categorical-categorical interactions are easier to interpret: you basically just get a coefficient for each combination of each level in the two categorical predictors and all of these coefficients can be though of as intercepts.

## The data: Song 2020

Let's read the data.

```{r}
#| label: shallow

library(tidyverse)

shallow <- read_csv("data/song2020/shallow.csv") |> 
  filter(Critical_Filler == "Critical", RT > 0)
```

We will be looking at RTs and relation type in L1 and L2 participants. Let's plot the RTs.

```{r}
#| label: shallow-plot
shallow |> 
  ggplot(aes(Relation_type, RT)) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  stat_summary(fun.data = "mean_cl_boot", aes(colour = Relation_type)) +
  facet_grid(cols = vars(Group)) +
  theme(legend.position = "none")
```

The means marked by coloured points in the plot suggest there might be a smaller difference in RTs between constituent and non-constituent/unrelated in the L2 than in the L1 group.

An interesting question we can ask is how much smaller is the difference in the L1 group compared to the L2 group. We can fit a regression model and wrangle the posterior draws to answer that question.

## The model

We want to allow the model to estimate mean RTs for each combination of relation type (`Relation_type`) and group (`Group`). Both predictors are categorical. We can achieve this by including an interaction term between the two categorical predictors: `Relation_type:Group`.

When both predictors are categorical, you only need the interaction term when suppressing the intercept with `0 +`. The model will estimate a coefficient for each combination of levels in the two predictors.

```{r}
#| label: shallow-bm

shallow_bm <- brm(
  logRT ~ 0 + Relation_type:Group,
  family = gaussian,
  data = shallow,
  file = "data/cache/regression-interactions-catcat_shallow_bm",
  cores = 4,
  seed = 6237
)
```

Check out the summary of the model. You will find 6 regression coefficients: 3 relation types (Constituent, Non-constituent, Unrelated) * 2 groups (L1 and L2).

```{r}
#| label: shallow-bm-summ
summary(shallow_bm, prob = 0.8)
```

We can easily plot this with `conditional_effects()`, like so.

```{r}
#| label: shallow-bm-cond
conditional_effects(shallow_bm, "Group:Relation_type")
```

Inspection of the estimated RTs for each relation type in the two groups suggests the difference between constituent and non-constituent/unrelated is larger in L1 than in L2 participants. Let's calculate the difference in the two groups and then the difference between these differences.

## Get the difference of differences

As usual, we need to obtain the posterior draws.

```{r}
#| label: shallo-bm-draws
shallow_bm_draws <- as_draws_df(shallow_bm)
```

Now, to get the difference between constituent and non-constituent/unrelated:

1. We must first get the mean of the non-constituent and unrelated relation types.
2. Then we can take the difference between the exponentiated posterior draws of the constituent type and the exponentiated posterior draws of the mean non-constituent/unrelated type. The exponentiation transforms the draws from logged ms to ms and then we take the difference.
3. Finally, we take the difference between the calculated differences in ms.

Run the code and inspect the `shallow_bm_diff` tibble to understand what is going on.

```{r}
#| label: shallow-bm-diff
shallow_bm_diff <- shallow_bm_draws |> 
  mutate(
    # 1. Mean non-constituent/unrelated for L1 and L2
    L1_NCU = mean(c(`b_Relation_typeNonConstituent:GroupL1`, `b_Relation_typeUnrelated:GroupL1`)),
    L2_NCU = mean(c(`b_Relation_typeNonConstituent:GroupL2`, `b_Relation_typeUnrelated:GroupL2`)),
    
    # 2. Difference between non-constituent/unrelated and constituent
    L1_diff = exp(L1_NCU) - exp(`b_Relation_typeConstituent:GroupL1`),
    L2_diff = exp(L2_NCU) - exp(`b_Relation_typeConstituent:GroupL2`),
    
    # 3. Difference of difference
    L2_L1_diff = L2_diff - L1_diff
  )
shallow_bm_diff |> 
  select(L1_NCU:L2_L1_diff)
```

The column `L2_L1_diff` contains the posterior draws of the difference between the constituent and non-constituent/unrelated type in the two groups. We can proceed as usual, like with any other posterior, and calculate the CrI of this difference of differences.

```{r}
#| label: shallow-bm-diff-post
library(posterior)

shallow_bm_diff |> 
  summarise(
    lo_80 = quantile2(L2_L1_diff, prob = 0.1),
    hi_80 = quantile2(L2_L1_diff, prob = 0.9),
    lo_80 = round(lo_80), hi_80 = round(hi_80),
  )
```

Here, at 80% confidence, the difference between constituent and non-constituent/unrelated types in the L2 group is 6 to 53 ms smaller than the difference in the L1 group.
