---
title: "Reproducibility and replicability"
description: "Learn about the meaning of reproducible, replicable, robust and generalisable research"
author: "Stefano Coretta"
date: 2024-09-15
format: 
  html:
    css: [webex.css]
    include-after-body: [webex.js]
execute: 
  warning: false
---

```{r}
#| label: setup
#| echo: false
#| message: false

library(tidyverse)
library(webexercises)
```

::: callout-warning
#### Prerequisites

-   [Research questions and hypotheses](research-qh.qmd)
:::

[![](../img/reproducible-definition-grid.svg){fig-align="center" width="550"}](https://doi.org/10.5281/zenodo.3332807)

## Definitions

The terms *reproducible* and *replicable* are often used interchangeably but The Turning Way has proposed a semantic scheme which introduces two more terms: *robust* and *generalisable*.

Check out the section [Table of Definitions for Reproducibility](https://book.the-turing-way.org/reproducible-research/overview/overview-definitions#table-of-definitions-for-reproducibility) from [The Turing Way book](https://book.the-turing-way.org), which defines the four terms.

::: callout-note
#### Quiz 1

```{r}
#| label: quiz-1
#| results: asis
#| echo: false

opts_1 <- c(
  "a. When the results do not match the researcher's expectations.",
  answer = "b. When the the same data and code as in the original study do not produce the published results.",
  "c. When research conducted by a different research team with new data does not produce the results as published in the original study. "
)

cat("**When is research not reproducible?**", longmcq(opts_1))
```

::: {.callout-tip collapse="true"}
#### Hint

Research is **reproducible** when you can produce the same results using the original data and code/methods.

Research is **replicable** when you can produce the same results using new data and the original code/methods.

See <https://the-turing-way.netlify.app/reproducible-research/overview/overview-definitions.html#table-of-definitions-for-reproducibility>.
:::
:::
